Cultural Transmission Can Explain the Evolution of Cooperative Behavior

### 2020-07-13 ms/main.tex
revised eq 20 and on

The frequency of $A$ among parents follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{table:1} and $\bar{w} = \hat{p} w_A + (1-\hat{p}) w_B = 1 + \hat{p}(b-c)$ is the population mean fitness.
Eq~\ref{eq:nextgen_parents} can be slimpified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

@@

In this case $0<v<1$, and the equation system is more complex than before.
Therefore, we focus on local rather than global stability, which was the case so far.\\
To proceed, we note that 
eq.~\ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
eq.~\ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\hat{p}$, and 
eq.~\ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we will find an equation for $\hat{p}'$ as a function of $\hat{p}$.
We will then determine the equilibria of this function, that is, solutions for $\hat{p}' = \hat{p}$, and their stability.

@@

For simplicity, assume that $T_A = T_B = T$ so we will get
\begin{equation} \label{eq:equal_horizontal_transmission}
  \hat{p}' - \hat{p} =  \hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T))
\end{equation}

We are looking for the equilibrium $\hat{p}^*$ which occurs when $\hat{p}' = \hat{p}$.
It is easy to verify that fixation of either phenotype is an equilibrium, that is,
\begin{equation} \label{eq:result2.1}
  \hat{p}^*_0 =  0, \, \hat{p}^*_1 = 1 .
\end{equation}
We wish to determine when fixation of cooperation is locally stable. 
An equilibrium point $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p}) = \hat{p}'-\hat{p}$ at $p^*$ is negative, $f'(\hat{p}^*)<0$. 
Calculating the derivative of $f(\hat{p})$ give us 
\begin{equation} \label{eq:derivative_of_phattag-phat}
 f'(\hat{p}) = 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
\end{equation}
For $\hat{p} = \hat{p}^*_1=1$, the derivative is
$f'(1) = -\alpha bvT + cv(1-T)$, and therefore, fixation of cooperation is locally stable when
\begin{equation} \label{eq:oblique_and_vertic_result2}
  \frac{b}{c}>\frac{1-T}{\alpha T}
\end{equation}
Which is the same condition we get in corollary 1.1.

@@

We now study the general case where $T_A \neq T_B$. We have
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \hat{p}'-\hat{p} = \beta_1p^3 + \beta_2 p^2 + \beta_3p + \beta_4
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{split}
  & \beta_1 = [c(1-v) -b (1-\alpha v)] \cdot(T_A-T_B) \\
  & \beta_2 = (b(1-\alpha v)-c(1-v)-1)(T_A-T_B)-\alpha bv T_A + cv(1-T_B) \\
  & \beta_3 = \alpha bvT_A - cv(1-T_B) + (T_A-T_B) \\
  & \beta_4 = 0
\end{split}
\end{equation}
Since $f(\hat{p})$ is a cubic polynomial, three equilibria exist.
We already know that $\hat{p}^*_0 = 0 $ and $\hat{p}^*_1 = 1$ are equilibria.
The third equilibrium can be found by solving $f(\hat{p})/\big(\hat{p}(1-\hat{p})\big)=0$, and it is
\begin{equation} \label{eq:oblique_and_vertic_result}
  \hat{p}^*_2 =  \frac{\alpha bvT_A + cvT_B + T_A-T_B-cv}{c(1-v)\cdot(T_A-T_B) -b (1-\alpha v) \cdot(T_A-T_B)}
\end{equation}

We would like to know when $\hat{p}^*_2$ is legitimate equilibrium, i.e. $0<\hat{p}^*_2<1$.  
To do that we find the derivative  of $f(\hat{p})$,
\begin{equation} \label{eq:derivative_of_phattag-phat2}
  f'(\hat{p}) = 3\beta_0 \hat{p}^2 + 2\beta_1\hat{p}^2 + \beta_3
 \end{equation}
We assume that $T_A>T_B$ and we look on $\beta_1$:
\begin{equation} \label{eq:beta1}
  \begin{split}
   \beta_1 & = [c(1-v) -b (1-\alpha v)] \cdot(T_A-T_B) \\
   &< [c(1-\alpha v) - b(1-\alpha v)]\cdot(T_A-T_B) \\
   &= (1-\alpha v)(c-b)(T_A-T_B)
  \end{split}
 \end{equation}
From eq.~\ref{eq:beta1} we get that $\beta_1<0$ since $c-b<0$ $1-\alpha v>0$ and we assumed that $T_A > T_B$.
In this case the cubic $f(\hat{p})$ goes to infinity when $\hat{p}$ goes to minus infinity ($-\infty$). 
Therefore, if $\hat{p}^*_2$ is positive then $f'(0)$ must be negative % are you sure? I think it must be positive, because i think that the two extremes of a cubic must go in opposite directions DC: Yest, but I've fixed in the line before, there was minus sign missing
\begin{equation} \label{eq:derivative_of_phattag-phat_p=0}
  f'(0) = \beta_3
 \end{equation}
$\hat{p}^*_2>0$ when $\beta_3<0$.
In addition, $\hat{p}^*_0=0$ is stable when $f'(0) = \beta_3 <0$.
Therefore, when $\hat{p}^*_2>0$ the coefficient $\beta_3$ must be negative and $\hat{p}^*_0=0$ is locally stable.

Now we would like to find when $\hat{p}^*_2<1$.
If $T_A>T_B$, as before, then the cubic $f(\hat{p})$ goes to minus infinity ($-\infty$) when $\hat{p}$ goes to infinity.
Hence, if $\hat{p}^*_2<1$ then $f'(1)$ must be negative % again, i think that the two extremes of a cubic must go in opposite directions. DC: I've double checked and I'm pretty sure my results are correct 
\begin{equation} \label{eq:derivative_of_phattag-phat_p=1}
  f'(1) = \beta_1 - \beta_3
\end{equation}
Therefore, $\hat{p}^*_2<1$ when $\beta_1-\beta_3<0$.
In addition, $\hat{p}^*_1=1$ is stable when $f'(1) = \beta_1-\beta_3<0$. 
We saw that when $0<\hat{p}^*_2<1$ the following must apply $f'(0)=\beta_3<0$ and $f'(1)=\beta_1-\beta_3<0$.
Hence, when $0<\hat{p}^*_2<1$ then $\hat{p}^*_1=1$ is locally stable.

\textbf{Result 4.1}: if $\hat{p}^*_2$ is a legitimate equilibrium, then it is locally unstable.

\textbf{Result 4.2}: if $T_A>T_B$, then $\hat{p}^*_2$ is a legitimate equilibrium when $\beta_1<\beta_3<0$.

\textbf{Result 4.3}: if $T_B>T_A$, then $\hat{p}^*_2$ is a legitimate equilibrium when $\beta_1>\beta_3>0$.

\begin{figure}[H]
  \centering
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2a.pdf}
    \caption{$V=1$, $T_A=T_B=T$, $\alpha \neq 0$}
    \label{fig:results_a}
  \end{subfigure}
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2b.pdf}
    \caption{$V=1$, $\alpha = 0$}
    \label{fig:results_b}
  \end{subfigure}
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2c.pdf}
    \caption{$V=0$}
    \label{fig:results_c}
  \end{subfigure}
  \label{fig:results}
  \caption{numerical results}
\end{figure}

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Ayelet Shavit for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR).
% TODO funding for other authors?
}

>>>>

The frequency of $A$ among parents follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{table:1}, and 
\begin{equation} \label{eq:mean_fitness}
\bar{w} = \hat{p} w_A + (1-\hat{p}) w_B = 1 + \hat{p}(b-c)
\end{equation}
is the population mean fitness.
Eq~\ref{eq:nextgen_parents} can be slimpified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

@@

In this case $0<v<1$, and the equation system is more complex than before.
Therefore, we focus on local rather than global stability, which was the case so far.\\
To proceed, we note that 
eq.~\ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
eq.~\ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\hat{p}$, and 
eq.~\ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$.
We then determine the equilibria of this equation, that is, solutions for $\hat{p}' = \hat{p}$, and their stability.

@@

For simplicity, we start with the assumption that $T_A = T_B = T$. We have
\begin{equation} \label{eq:equal_horizontal_transmission}
% DOR: I checked again, substituting eq 7 and 9 to eq 5: you have to divide the RHS by wbar.
  %\hat{p}' - \hat{p} = 
  %\hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T)) = 
  \bar{w}(\hat{p}' - \hat{p}) = 
  f(\hat{p}) = \hat{p}(1-\hat{p})(\alpha bvT - cv(1-T)) % shorter version
\end{equation}

We are looking for the equilibrium $\hat{p}^*$ which occurs when $f(\hat{p}^*)=0$, or $\hat{p}' = \hat{p}$.
It is easy to verify that fixation of either phenotype is an equilibrium, that is,
\begin{equation} \label{eq:result2.1}
  \hat{p}^*_0 =  0, \, \hat{p}^*_1 = 1 .
\end{equation}

Is fixation of cooperation, $\hat{p}^*_1=1$, locally stable?. 
An equilibrium point $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})$ at $p^*$ is negative, $f'(\hat{p}^*)<0$. 
This derivative is
\begin{equation} \label{eq:derivative_of_phattag-phat}
	f'(\hat{p}) \mid_{\hat{p}=1} = 
% 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
	(1- 2\hat{p}) (\alpha bvT - cv(1-T)) \mid_{\hat{p}=1} = 
	-\alpha bvT + cv(1-T)).
\end{equation}
Therefore, fixation of cooperation is locally stable when
\begin{equation} \label{eq:oblique_and_vertic_result2}
  %\frac{b}{c}>\frac{1-T}{\alpha T}
	c < b \cdot \frac{\alpha T}{1-T},
\end{equation}
which is the same condition we found in Corollary 1.1, Eq.~\ref{eq:equal_transmission}.

@@

We now study the general case where $T_A \neq T_B$. We have
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) \\
%\beta_2 &= -[c(1-v) - b(1-\alpha v) +  1](T_A-T_B)-\alpha bv T_A + cv(1-T_B) \\
\beta_2 &= -\beta_1 -\beta_3  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}
Since $f(\hat{p})$ is a cubic polynomial, three equilibria exist.
We already know that $\hat{p}^*_0 = 0 $ and $\hat{p}^*_1 = 1$ are equilibria.
The third equilibrium $\hat{p}^*_2$ can be found by solving $f(\hat{p})/\big[\hat{p}(1-\hat{p})\big] = \beta_3 -\beta_1 \hat{p} = 0$:
\begin{equation} \label{eq:oblique_and_vertic_result}
  \hat{p}^*_2 =  
  \frac{\beta_3}{\beta_1}.
  %\frac{\alpha bvT_A - cv(1-T_B) + (T_A-T_B)}
  %     {\big(c(1-v) -b (1-\alpha v)\big) (T_A-T_B)} .
\end{equation}
To determine when $\hat{p}^*_2$ is a legitimate equilibrium (i.e. $0<\hat{p}^*_2<1$), we first focus on the left inequality, $0<\hat{p}^*_2$, and
%To do that we find the derivative  of $f(\hat{p})$,
%\begin{equation} \label{eq:derivative_of_phattag-phat2}
%  f'(\hat{p}) = 3\beta_1 \hat{p}^2 + 2\beta_2\hat{p}^2 + \beta_3
% \end{equation}
assume that $T_A>T_B$. In this case,
\begin{equation} \label{eq:beta1}
  \begin{split}
   \beta_1 & = [c(1-v) - b (1-\alpha v)] (T_A-T_B) \\
   &< [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) \\
   &= (1-\alpha v)(c-b)(T_A-T_B) < 0 ,
  \end{split}
 \end{equation}
since $T_A > T_B$, $c<b$, and $1>\alpha v$.
Therefore, when $\hat{p}$ goes to negative infinity, the cubic $f(\hat{p})$ goes to positive infinity.
So, a necessary condition for $\hat{p}^*_2>0$ is that $f'(0) = \beta_3 < 0$.
This condition also satisfies local stability of $\hat{p}^*_0=0$.
Hence, if $\hat{p}^*_2$ is legitimate, then extinction of cooperation is locally stable.

We now focus on $\hat{p}^*_2<1$, still assuming $T_A>T_B$.
When $\hat{p}$ goes to positive infinity, the cubic $f(\hat{p})$ goes to negative infinity (Eq.~\ref{eq:beta1}).
So, a necessary condition for $\hat{p}^*_2<1$ is that $f'(1)=\beta_1 - \beta_3<0$.
This condition also satisfies local stability of $\hat{p}^*_1=1$.
Hence, if $\hat{p}^*_2$ is legitimate, then fixation of cooperation is also locally stable.
Moreover, since legitimacy of $\hat{p}^*_2$ implies local stability of both $\hat{p}^*_0$ and $\hat{p}^*_1$, we conclude that $\hat{p}^*_2$ is locally unstable.

If we assume instead that $T_A < T_B$, we reach the following results:

\textbf{Result 4}: if $T_A>T_B$ ($T_B>T_A$), then $\hat{p}^*_2$ is a legitimate locally unstable equilibrium when $\beta_1<\beta_3<0$ ($\beta_1>\beta_3>0$). In this case, both fixations are locally stable and $\hat{p}^*_2$ separates the domains of attraction.

\begin{figure}[H]
  \centering
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2a.pdf}
    \caption{$v=1$, $T_A=T_B=T$, $\alpha \neq 0$}
    \label{fig:results_a}
  \end{subfigure}
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2b.pdf}
    \caption{$v=1$, $\alpha = 0$}
    \label{fig:results_b}
  \end{subfigure}
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2c.pdf}
    \caption{$v=0$}
    \label{fig:results_c}
  \end{subfigure}
  \label{fig:results}
  \caption{
  \textbf{Numerical results for cultural evolution of cooperation.}
  Shown are dynamics of \textbf{(a-b)} $\tilde{p}$, the frequency of parents with cooperative phenotype $A$; \textbf{(c)} $p'$, the frequency of adults with cooperative phenotype $A$.
  The figure demonstrates fixation of cooperation (green), extinction of cooperation (blue)m and stable co-existence of cooperators and defectors (orange). %TODO ic (c), the blue and green got mixed. 
  % TODO have all panels in one row.
  }
\end{figure}

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Ayelet Shavit for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR),
and Minerva Center for Lab Evolution (YR).
% TODO funding for other authors?
}
